\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-04-14 forwards-and-backwards.tex
edits with alison

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. The vast majority of these
results concern spin systems \comment{REFs}. However,~\cite{flandoli/huang:2021} 
obtain the Fisher-KPP equation
as a scaling limit of interacting diffusions in
$\IR^d$. In that setting a major challenge (also apparent in our models), not
faced in spin systems, is 
the control of concentration of particles. We also note that the scaling 
considered in~\cite{flandoli/huang:2021} is different from that considered
below. In their language we consider only `moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas they also consider the situation in which
that number remains finite;
furthermore we shall consider a scaling in which 
the reproduction mechanism is near-critical and look at large timescales,
while \citet{flandoli/huang:2021} only scale the number of
particles in the system and the size of the neighbourhood over which indivduals interact.

@@

\comment{This paragraph to be rewritten last}

@@

@@

\begin{figure}
    \begin{center}
        \includegraphics{figures/ex1a/fkpp_123.density.3.0}
        \includegraphics{figures/ex1b/fkpp_123.density.3.0}
    \end{center}
    \caption{
        Snapshots of two simulations, with small $\alpha=\theta/N$ (left) and large 
	$\alpha =\theta/N$ (right).
        Simulations are run with a Fisher-KPP-like parameterization:
        birth and establishment are constant, while death increases linearly with density,
        at slope $1/\theta$.
        Left: $\alpha=0.1$. Right: $\alpha=10$.
        Other parameters were the same:
        dispersal and interactions distance were set to 1,
        and the equilibrium density is 10 individuals per unit area.
        Blue curves are contours from a kernel density estimate.
        \label{fig:super_vs_det_2d}
    }
\end{figure}

@@

\comment{I'm a bit surprised by the factor of 2 (that manifests as $\varphi^2$),
    but I've checked where it comes from pretty carefully, and don't see anything wrong.}
Long-term fitness of an individual
is proportional to the fraction of lineages from the distant future that pass through the individual,
and hence the total long-term fitness at a location is proportional to
the stationary distribution of $Y$ there, if it exists.
Therefore,
if $\pi$ is integrable then the per-capita long-term fitness of an individual at $x$
is proportional to $\pi(x) / \varphi(x)$.

@@

\comment{Maybe check for more recent references for paragraph below.}
Equation~(\ref{eqn:nonlocal_pde}) is a nonlocal version of a reaction diffusion equation with nonlinear
diffusion; in other words the diffusivity of the population depends on the population
density. Passing to the classical limit, we recover equations
like~(\ref{PME}).
Such equations are widely used in a number of contexts in biology in which
it is clear that motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see~\cite{sherratt:2010} and references therein for further discussion. 
In particular, it has been suggested as a model for
the expansion of a certain type of bacteria % of the type {\em Paenbacillus dendritiformis}
on a thin layer of agar in a petri dish 
\citep{cohen/golding/kozlovsky/benjacob/ron:1999}. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population. 
\cite{li/buenzli/simpson:2022} write down an interacting particle
system on a hexagonal
lattice (with at most one individual per site) that incorporates an effect
of local crowding on individual motion. Their equation
corresponds to a reaction-diffusion equation with nonlinear diffusion and a 
bistable reaction term of the form~(\ref{Birzu nonlinearity}) and they
investigate the interplay between nonlinear diffusion and long-term 
survival or extinction of the population. We focus instead on 
the monostable reaction of~(\ref{PME}).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/ex2a/pme_123.steps}
        \includegraphics[width=0.45\textwidth]{figures/ex2b/pme_123.steps}
    \end{center}
    \caption{
        Two panels, showing simulated 1D populations under a
porous medium equation with logistic growth~(\ref{PME}), comparing the simulated
        profile to the analytical solution;
        $\theta/N$ not small on left; small on right.
        % (Note: noisier wave should move slower, although maybe not much since it's pushed.)
        \label{fig:pme_waves}
    }
\end{figure}

@@

The implication here is that the equilibrium solution to~\eqref{general deterministic limit}
may not be unique,
and a stable solution may have oscillations on a scale determined by the interaction distance.
We can see this easily in the simple case that
$\DG = \sigma^2 \Delta$,
the interaction kernels $\rho_r$, $\rho_\gamma$, and $\rho_F$
are all equal to $p_{\epsilon^2}$ (the heat kernel at time $\epsilon^2$),
and all parameters are spatially homogeneous,
so that $r(x, \eta) = r(\smooth{r}\eta(x))$, and similarly for $\gamma$ and $F$.
Let $\varphi_0$ be a value so that $F(\varphi_0) = 0$ and $F'(\varphi_0) < 0$,
so that we would expect $\varphi(x) \equiv \varphi_0$ to be a stable equilibrium solution.
Let us linearize~\eqref{general deterministic limit} about $\varphi_0$:
let $\varphi_t(x) = \varphi_0 + \psi_t(x)$,
and (informally) $r(x) \approx r(\varphi_0) + r'(\varphi_0) \smooth{r} \psi(x)$.
Writing $r_0 = r(\varphi_0)$ and $r'_0 = r'(\varphi_0)$,
and similarly for $\gamma$ and $F$,
\begin{align*}
    \partial_t \psi
    &\approx
    \sigma^2 \varphi_0 r_0 \gamma_0' \Delta \smooth{\gamma} \psi
    + \sigma^2 r_0 \gamma_0 \Delta \psi
    + \varphi_0 F'_0 \smooth{F} \psi .
\end{align*}
Letting $\hat \psi(u) = \int e^{2\pi i u x} \psi(x) dx$,
\begin{align*}
    \partial_t \hat \psi(u)
    &\approx
    \left(
        - \sigma^2 \varphi_0 r_0 \gamma_0' u^2 e^{-\epsilon^2 u^2 /2}
        - \sigma^2 r_0 \gamma_0 u^2
        + \varphi_0 F'_0 e^{-\epsilon^2 u^2 /2} 
    \right) \hat \psi(u) .
\end{align*}
Since $F'_0 < 0$, the last two terms are negative,
but on the other hand if we take $\gamma_0' < 0$
(i.e., per capita birth rate decreases with density near equilibrium),
then the stability of the constant solution $\varphi \equiv \varphi_0$
depends on the balance between the first and latter two terms.
Writing
$A_1 = - \sigma^2 \varphi_0 r_0 \gamma'_0$,
$A_2 = \sigma^2 r_0 \gamma_0$,
and $A_3 = - \varphi_0 F'_0$
(with signs chosen so each is positive),
and defining
$$ g(v) = A_1 v - A_2 v e^{\epsilon^2 v /2} - A_3 , $$
we have $\partial_t \hat \psi(u) = g(u^2) e^{\epsilon^2 u^2 /2} \psi(u)$.
The term $g(v)$ is maximized at $v_*$ solving
$e^{\epsilon^2 v_*/2} = A_1 / (A_2 (\epsilon^2 v_*/ 2 + 1))$,
which is positive if $A_1 > A_2$,
i.e., if $-\varphi_0 \gamma_0' > \gamma_0$.
If so, then the constant solution is unstable with respect to perturbations
that oscillate with frequency around $\sqrt{v_*}$.
\comment{This seems wrong, since it doesn't involve $\sigma$?
    But at least note the unstable frequencies scale with $\epsilon$.
    We might say more by linearizing the exponential term,
    which gives $v_* = (A_1 - A_2 + A_3 \epsilon^2/2)/(A_1 \epsilon^2)$,
    although the resulting expression isn't terribly helpful.
}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/clumpy_2d/clumpy_123.density.3.0}
        \includegraphics[width=0.45\textwidth]{figures/clumpy_1d/clumpy_123.steps}
    \end{center}
    \caption{
        \textbf{Left:} A snapshot of individual locations in a two-dimensional simulation
        in which the constant density is unstable
        and a stable, periodic pattern forms.
        \textbf{Right:} Population density in an expanding wave in a one-dimensional simulation
        forming a period pattern;
        each panel shows the wave front in three periods of time;
        within each period of time the wave front at earlier times is shown in blue
        and later times in pink.
        In both cases, $\gamma(m) = 3/(1 + m)$, $\mu \equiv 0.3$, and $r \equiv 1$;
        dispersal is Gaussian with $\sigma=0.2$ and density is measured with $\rho_\gamma(x) = p_9(x)$,
        i.e., using a Gaussian kernel with standard deviation 3.
        \label{fig:clumping}
    }
\end{figure}

@@

In the following sections we present formal proofs of our three scaling limits.
In Subsection~\ref{sec:population_density_proof} 
we prove Theorem~\ref{thm:nonlocal_convergence}, to obtain a limit in
which interactions between individuals in the population are nonlocal.
In Subsection~\ref{subsec:nonlocal to local}
we show how, in two important examples in which the nonlocal
limit is respectively a deterministic solution to a non-local equation of 
reaction diffusion type and a deterministic solution to a nonlocal 
porous medium equation with an additional logistic growth term, 
one can pass to a further 
limit to obtain a classical PDE.
In subsequent sections (Sections~\ref{subsec:one step convergence proof} and~\ref{sec:lookdown_proofs})
we prove our third scaling limit, in which 
we pass from our population model 
directly to a reaction diffusion equation without going via a nonlocal limit,
and finally our results concerning both convergence of the lookdown representation
and characterization of the lineage process in the deterministic case.
\comment{TODO: revisit this paragraph}

@@

\begin{lemma}[Tightness of $(\langle f, \eta^{N}_t \rangle )_{t>0}$]
    \label{lem:eta_projections_tightness}
	For each $f \in C^{\infty}_{b}(\overline{\IR}^d)$, 
the collection of processes
$(\langle f, \eta^{N}_t \rangle)_{t \geq 0}$
for $N = 1, 2, \ldots$ is tight
as a sequence of c\`adl\`ag, real-valued processes.
\end{lemma}
\begin{proof}
The Aldous-Rebolledo criterion (found in the form we use here in \cite{rebolledo:1980}),
applied to the semimartingale representation of $\langle f, \eta^N_t\rangle$
of equation~\eqref{eqn:eta_f_mgale_decomp}, tells us
that it suffices to show that for each $T>0$,
(a) for each fixed $0\leq t\leq T$, the sequence $\{\langle f, \eta^N_t \rangle\}_{N \ge 1}$ is tight,
and (b) for any sequence of stopping times $\tau_N$ bounded by $T$,
and for each $\nu > 0$, there exist $\delta > 0$ and $N_0 > 0$ such that 
\begin{gather}
        \label{eqn:eta_projections_goal1}
    \sup_{N > N_0}
    \sup_{t \in [0, \delta]}
    \IP\left\{\left|
            \int_\tau^{\tau + t}
            \int_{\IR^d}
            \left\{
                \gamma(x, \eta^N_s) B^\theta_f(x, \eta^N_s)
                + f(x) F(x, \eta^N_s)
            \right\} 
            \eta^N_s(dx)
            ds
        \right|> \nu \right\}
        < \nu ,
    \\ \text{and} \qquad
        \label{eqn:eta_projections_goal2}
    \sup_{N > N_0}
    \sup_{t \in [0, \delta]}
    \IP\left\{\big|
        [M^{N}(f)]_{\tau + t} 
            - [M^{N}(f)]_\tau \big|
        > \nu
    \right\}
    < \nu.
\end{gather}
Tightness of $\langle f, \eta^N_t\rangle$ for fixed $t$
follows from Lemma~\ref{lem:eta_f_bound} and Markov's inequality,
so we focus on the remaining conditions.

@@

\begin{lemma}
\label{no mass at infty}
Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
let $(\eta_t)_{t\geq 0}$ be a limit point of $\{(\eta^N_t)_{t\geq 0}\}_{N\geq 1}$. 
For any $\delta>0$, 
\[
\IP\big[\eta_t\big(\{\|x\|>R\}\big)>\delta\big]\to 0\qquad\mbox{ as }R\to\infty.
\]
\end{lemma}
\begin{proof}[Sketch]
To avoid the idea being buried in notation, we sketch the proof in the 
special case of $d=1$ and $\DG=\Delta$.
For any $f\in C_b^\infty(\IR)$,
\begin{multline}
\label{special case first moment equation}
\IE[\langle f,\eta_t\rangle] 
=\IE[\langle f,\eta_0\rangle]
\\
+\IE\Big[\int_0^t\Big\langle\gamma(x, \smooth{\gamma}\eta_s(x))
\Delta \big(f(\cdot)r(\cdot,\smooth{r}\eta_s(\cdot))\big)(x)
+F(x,\smooth{F}\eta_s(x)) f(x),
\eta_s(dx)\Big\rangle ds\Big].
\end{multline} 
We take a sequence of test functions $f_n$ increasing to
$\exp(\lambda x)$, with $\lambda>0$. Having passed
to the limit,
substituted in~(\ref{special case first moment equation}) and 
expanded the term involving the Laplacian, we obtain
\begin{multline}
\label{exponential moment}
    \IE[\langle \exp(\lambda x),\eta_t\rangle]
    =
    \IE[\langle \exp(\lambda x),\eta_0\rangle]
    +
    \IE\Big[\int_0^t\Big\langle\gamma(x, \smooth{\gamma}\eta_s(x))
    \exp(\lambda x)
    \Big\{
        \lambda^2 r(x, \smooth{r}\eta_s(x))
    \\ {}
        + 2\lambda\partial_x\big(r(\cdot,\smooth{r}\eta_s(\cdot)\big)(x)
        + \partial_{xx} r(\cdot,\smooth{r}\eta_s(\cdot)\big)(x)
    \Big\} 
    + F(x,\smooth{F}\eta_s(x)),
    \eta_s(dx)\Big\rangle ds\Big] .
\end{multline} 
    \comment{alison to look at this again; check whether it depends on just bounding $B^\theta_1$}
Evidently, in much the same way as in the proof of
Lemma~\ref{lem:eta_f_bound},
our conditions on $r$ and $F$ ensure
that the integrand on the right hand side 
of~(\ref{exponential moment})
is bounded by $C\langle\exp(\lambda x),\eta_s(dx)\rangle$ for some constant
$C$. Gronwall's inequality then implies that 
$\IE[\langle\exp(\lambda x),\eta_t(dx)\rangle]$ 
is uniformly bounded for $t\in [0,T]$, and an application of Markov's inequality
tells us that for any $\delta >0$,
$\IP[\eta_t([R,\infty])>\delta]\to 0$ as $R\to\infty$. A symmetric argument
yields the corresponding bound for $\IP[\eta_t([-\infty, -R])>\delta$ and the 
result is proved.
\end{proof}

@@

\comment{below $\eta_s^N$ stops having  an $N$ at some point}

@@

\comment{perhaps explain: the next proposition is useful because we'll want to use weak convergence of $\smooth{\epsilon}\eta_s$; but in the martingale we have the term on the left in the proposition; probably there are other ways to do this.}
\comment{ACTUALLY BETTER we want to transfer $T_\epsilon^2$ over onto $\eta$; this happens in two steps, first this proposition; next using that $T_{\epsilon^2}f - f$ is small}

@@

Combining the above, we have reduced the problem to showing that 
\begin{equation}
\label{target expression}
	\lim_{\epsilon\to 0}
	\int_0^t\big\langle \int |\rho_\epsilon*\eta_s(y)-\rho_\epsilon*\eta_s(w)|
\left(\rho_\epsilon*\eta_s(y))^k+\rho_\epsilon*\eta_s(w)^k\right)
p_{\epsilon^2}(w,y)dy, \eta_s(dw)\big\rangle ds
=0,
\end{equation}
\comment{which we prove by showing the expectation of this positive quantity goes to zero; e.g. see Lemma 7.5(?)}.

@@

\comment{Probably should put a bit more detail in here?}
\comment{key bit is that we can always write $M_s(w)$ as a convolution of $p_{\epsilon^2/}$ with something}

@@

    The analogue of Lemma~\ref{lem:limit_mgale},
    in particular that the martingale properties are preserved under passage to the limit,
    are proved in Section~\ref{sec:lookdown_generator_proofs}.
    \comment{TODO: refer to correct thing}
    Finally, we must show that the limiting lookdown process $\lp$
    projects to the limiting process $\eta$, i.e., a solution of the martingale
    problem in Theorem~\ref{thm:nonlocal_convergence}.
    Let $N_k \to \infty$ be a sequence along which $\lp^{N_k}$ converges.
    By Theorem~\ref{thm:nonlocal_convergence},
    there is a subsequence $N_{k(j)}$ along which the projected population processes
    $\eta^{N_{k(j)}}$ converge, and the limit solves the martingale problem.
    Thus any limit point of $\lp^N$ projects to a population process $\eta$
    solving the martingale problem of Theorem~\ref{thm:nonlocal_convergence}.
\end{proof}

@@

We postpone the proof of this Lemma to show how it gets us to compact containment.
First, we show that this implies compact containment for arbitrary
compactly supported functions.

@@

In other words, the processes $\lp^N$ stay in 
the set
$$
    \left\{
        \lp \in \lpmeasures \;:\;
        \langle h_k, \lp \rangle \le C_k
        \text{ for all } k \ge 1
    \right\} ,
$$
for all $0 \le t \le T$
with uniformly high probability,
a set which is compact if we choose $\{f_k\}$ appropriately.
% \comment{
%     Note that we could instead of the vague topology on locally finite measures,
%     take the space of measures with $\langle h, \xi \rangle < \infty$
%     for which $\xi_n \to \xi$ if $\langle f, \xi \rangle < \infty$ for all $f$ for which $f \le c_f h$,
%     which is completely metrizable
%     as in Condition 2.1 of \citet{etheridge/kurtz:2019}.
% }

>>>>

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. The vast majority of these
results concern spin systems \comment{TODO REFs obtained from Flandoli/Huang etc: Alison}.
However,~\cite{flandoli/huang:2021} 
obtain the Fisher-KPP equation
as a scaling limit of interacting diffusions in
$\IR^d$. In that setting a major challenge (also apparent in our models), not
faced in spin systems, is 
the control of concentration of particles. We also note that the scaling 
considered in~\cite{flandoli/huang:2021} is different from that considered
below. In their language we consider only `moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas they also consider the situation in which
that number remains finite;
furthermore we shall consider a scaling in which 
the reproduction mechanism is near-critical and look at large timescales,
while \citet{flandoli/huang:2021} only scale the number of
particles in the system and the size of the neighbourhood over which indivduals interact.

@@

\comment{TODO: This paragraph to be rewritten last PETER}

@@

\comment{TODO: modify condition (4) to say that the $C_f$ only depends on $\|f\|_\infty$, $\|\partial f\|$ and $\|\partial^2 f\|$: PETER}

@@

\begin{figure}
    \begin{center}
        \includegraphics{figures/ex1a/fkpp_123.density.3.0}
        \includegraphics{figures/ex1b/fkpp_123.density.3.0}
    \end{center}
    \caption{
        Snapshots of two simulations, with small $\alpha=\theta/N$ (left) and large 
	$\alpha =\theta/N$ (right).
        Simulations are run with a Fisher-KPP-like parameterization:
        birth and establishment are constant, while death increases linearly with density,
        at slope $1/\theta$.
        Left: $\alpha=0.1$. Right: $\alpha=10$.
        Other parameters were the same:
        dispersal and interactions distance were set to 1,
        and the equilibrium density is 10 individuals per unit area.
        \label{fig:super_vs_det_2d}
    }
\end{figure}

@@

% \comment{I'm a bit surprised by the factor of 2 (that manifests as $\varphi^2$),
%     but I've checked where it comes from pretty carefully, and don't see anything wrong.}
Long-term fitness of an individual
is proportional to the fraction of lineages from the distant future that pass through the individual,
and hence the total long-term fitness at a location is proportional to
the stationary distribution of $Y$ there, if it exists.
Therefore,
if $\pi$ is integrable then the per-capita long-term fitness of an individual at $x$
is proportional to $\pi(x) / \varphi(x)$.

@@

Equation~(\ref{eqn:nonlocal_pde}) is a nonlocal version of a reaction diffusion equation with nonlinear
diffusion; in other words the diffusivity of the population depends on the population
density. Passing to the classical limit, we recover equations
like~(\ref{PME}).
Such equations are widely used in a number of contexts in biology in which
it is clear that motility within a population varies with population density.
For example, density dependent dispersal is a common feature in spatial
models in ecology, eukaryotic cell biology, and avascular tumour growth;
see~\cite{sherratt:2010} and references therein for further discussion. 
In particular, it has been suggested as a model for
the expansion of a certain type of bacteria % of the type {\em Paenbacillus dendritiformis}
on a thin layer of agar in a petri dish 
\citep{cohen/golding/kozlovsky/benjacob/ron:1999}. 
We shall pay particular attention to the case in which the equation can be 
thought of as modelling the density of an expanding population. 
\cite{li/buenzli/simpson:2022} write down an interacting particle
system on a hexagonal
lattice (with at most one individual per site) that incorporates an effect
of local crowding on individual motion. Their equation
corresponds to a reaction-diffusion equation with nonlinear diffusion and a 
bistable reaction term of the form~(\ref{Birzu nonlinearity}) and they
investigate the interplay between nonlinear diffusion and long-term 
survival or extinction of the population. We focus instead on 
the monostable reaction of~(\ref{PME}).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/ex2a/pme_123.steps}
        \includegraphics[width=0.45\textwidth]{figures/ex2b/pme_123.steps}
    \end{center}
    \caption{
        \comment{TODO PETER; maybe remove dotted lines}
        Two panels, showing simulated 1D populations under a
porous medium equation with logistic growth~(\ref{PME}), comparing the simulated
        profile to the analytical solution;
        $\theta/N$ not small on left; small on right.
        % (Note: noisier wave should move slower, although maybe not much since it's pushed.)
        \label{fig:pme_waves}
    }
\end{figure}

@@

The implication here is that the equilibrium solution to~\eqref{general deterministic limit}
may not be unique,
and a stable solution may have oscillations on a scale determined by the interaction distance.
We can see this easily in the simple case that
$\DG = \sigma^2 \Delta$,
the interaction kernels $\rho_r$, $\rho_\gamma$, and $\rho_F$
are all equal to $p_{\epsilon^2}$ (the heat kernel at time $\epsilon^2$),
and all parameters are spatially homogeneous,
so that $r(x, \eta) = r(\smooth{r}\eta(x))$, and similarly for $\gamma$ and $F$.
Let $\varphi_0$ be a value so that $F(\varphi_0) = 0$ and $F'(\varphi_0) < 0$,
so that we would expect $\varphi(x) \equiv \varphi_0$ to be a stable equilibrium solution.
Let us linearize~\eqref{general deterministic limit} about $\varphi_0$:
let $\varphi_t(x) = \varphi_0 + \psi_t(x)$,
and (informally) $r(x) \approx r(\varphi_0) + r'(\varphi_0) \smooth{r} \psi(x)$.
Writing $r_0 = r(\varphi_0)$ and $r'_0 = r'(\varphi_0)$,
and similarly for $\gamma$ and $F$,
\begin{align*}
    \partial_t \psi
    &\approx
    \sigma^2 \varphi_0 r_0 \gamma_0' \Delta \smooth{\gamma} \psi
    + \sigma^2 r_0 \gamma_0 \Delta \psi
    + \varphi_0 F'_0 \smooth{F} \psi .
\end{align*}
Letting $\hat \psi(u) = \int e^{2\pi i u x} \psi(x) dx$,
\begin{align*}
    \partial_t \hat \psi(u)
    &\approx
    \left(
        - \sigma^2 \varphi_0 r_0 \gamma_0' u^2 e^{-\epsilon^2 u^2 /2}
        - \sigma^2 r_0 \gamma_0 u^2
        + \varphi_0 F'_0 e^{-\epsilon^2 u^2 /2} 
    \right) \hat \psi(u) .
\end{align*}
Since $F'_0 < 0$, the last two terms are negative,
but on the other hand if we take $\gamma_0' < 0$
(i.e., per capita birth rate decreases with density near equilibrium),
then the stability of the constant solution $\varphi \equiv \varphi_0$
depends on the balance between the first and latter two terms.
Writing
$A_1 = - \sigma^2 \varphi_0 r_0 \gamma'_0$,
$A_2 = \sigma^2 r_0 \gamma_0$,
and $A_3 = - \varphi_0 F'_0$
(with signs chosen so each is positive),
and defining
$$ g(v) = A_1 v - A_2 v e^{\epsilon^2 v /2} - A_3 , $$
we have $\partial_t \hat \psi(u) = g(u^2) e^{\epsilon^2 u^2 /2} \psi(u)$.
The term $g(v)$ is maximized at $v_*$ solving
$e^{\epsilon^2 v_*/2} = A_1 / (A_2 (\epsilon^2 v_*/ 2 + 1))$,
which is positive if $A_1 > A_2$,
i.e., if $-\varphi_0 \gamma_0' > \gamma_0$.
If so, then the constant solution is unstable with respect to perturbations
that oscillate with frequency around $\sqrt{v_*}$.
\comment{This seems wrong, since it doesn't involve $\sigma$?
    But at least note the unstable frequencies scale with $\epsilon$.
    We might say more by linearizing the exponential term,
    which gives $v_* = (A_1 - A_2 + A_3 \epsilon^2/2)/(A_1 \epsilon^2)$,
    although the resulting expression isn't terribly helpful.
    TODO ALISON
}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/clumpy_2d/clumpy_123.density.3.0}
        \includegraphics[width=0.45\textwidth]{figures/clumpy_1d/clumpy_123.steps}
    \end{center}
    \caption{
        \textbf{Left:} A snapshot of individual locations in a two-dimensional simulation
        in which the constant density is unstable
        and a stable, periodic pattern forms.
        \textbf{Right:} Population density in an expanding wave in a one-dimensional simulation
        forming a periodic pattern;
        each panel shows the wave front in three periods of time;
        within each period of time the wave front at earlier times is shown in blue
        and later times in pink.
        In both cases, $\gamma(m) = 3/(1 + m)$, $\mu \equiv 0.3$, and $r \equiv 1$;
        dispersal is Gaussian with $\sigma=0.2$ and density is measured with $\rho_\gamma(x) = p_9(x)$,
        i.e., using a Gaussian kernel with standard deviation 3.
        \label{fig:clumping}
    }
\end{figure}

@@

In the following sections we present formal proofs of our three scaling limits.
In Subsection~\ref{sec:population_density_proof} 
we prove Theorem~\ref{thm:nonlocal_convergence}, to obtain a limit in
which interactions between individuals in the population are nonlocal.
In Subsection~\ref{subsec:nonlocal to local}
we show how, in two important examples in which the nonlocal
limit is respectively a deterministic solution to a non-local equation of 
reaction diffusion type and a deterministic solution to a nonlocal 
porous medium equation with an additional logistic growth term, 
one can pass to a further 
limit to obtain a classical PDE.
In subsequent sections (Sections~\ref{subsec:one step convergence proof} and~\ref{sec:lookdown_proofs})
we prove our third scaling limit, in which 
we pass from our population model 
directly to a reaction diffusion equation without going via a nonlocal limit,
and finally our results concerning both convergence of the lookdown representation
and characterization of the lineage process in the deterministic case.
\comment{TODO: check/improve this paragraph: ALISON}

@@

\begin{lemma}[Tightness of $(\langle f, \eta^{N}_t \rangle )_{t>0}$]
    \label{lem:eta_projections_tightness}
	For each $f \in C^{\infty}_{b}(\overline{\IR}^d)$, 
the collection of processes
$(\langle f, \eta^{N}_t \rangle)_{t \geq 0}$
for $N = 1, 2, \ldots$ is tight
as a sequence of c\`adl\`ag, real-valued processes.
\end{lemma}
\begin{proof}
The Aldous-Rebolledo criterion (Theorem~\ref{thm:aldous_rebolledo})
applied to the semimartingale representation of $\langle f, \eta^N_t\rangle$
of equation~\eqref{eqn:eta_f_mgale_decomp}, tells us
that it suffices to show that for each $T>0$,
(a) for each fixed $0\leq t\leq T$, the sequence $\{\langle f, \eta^N_t \rangle\}_{N \ge 1}$ is tight,
and (b) for any sequence of stopping times $\tau_N$ bounded by $T$,
and for each $\nu > 0$, there exist $\delta > 0$ and $N_0 > 0$ such that 
\begin{gather}
        \label{eqn:eta_projections_goal1}
    \sup_{N > N_0}
    \sup_{t \in [0, \delta]}
    \IP\left\{\left|
            \int_\tau^{\tau + t}
            \int_{\IR^d}
            \left\{
                \gamma(x, \eta^N_s) B^\theta_f(x, \eta^N_s)
                + f(x) F(x, \eta^N_s)
            \right\} 
            \eta^N_s(dx)
            ds
        \right|> \nu \right\}
        < \nu ,
    \\ \text{and} \qquad
        \label{eqn:eta_projections_goal2}
    \sup_{N > N_0}
    \sup_{t \in [0, \delta]}
    \IP\left\{\big|
        [M^{N}(f)]_{\tau + t} 
            - [M^{N}(f)]_\tau \big|
        > \nu
    \right\}
    < \nu.
\end{gather}
Tightness of $\langle f, \eta^N_t\rangle$ for fixed $t$
follows from Lemma~\ref{lem:eta_f_bound} and Markov's inequality,
so we focus on the remaining conditions.

@@

\begin{lemma}
\label{no mass at infty}
Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
let $(\eta_t)_{t\geq 0}$ be a limit point of $\{(\eta^N_t)_{t\geq 0}\}_{N\geq 1}$. 
For any $\delta>0$, 
\[
\IP\big[\eta_t\big(\{\|x\|>R\}\big)>\delta\big]\to 0\qquad\mbox{ as }R\to\infty.
\]
\end{lemma}

\begin{proof}[Sketch]
    \comment{Change this to a sequence $f_n$ of smooth things with uniformly bounded derivs that increase 1; follows from condition 4. PETER}
To avoid the idea being buried in notation, we sketch the proof in the 
special case of $d=1$ and $\DG=\Delta$.
For any $f\in C_b^\infty(\IR)$,
\begin{multline}
\label{special case first moment equation}
\IE[\langle f,\eta_t\rangle] 
=\IE[\langle f,\eta_0\rangle]
\\
+\IE\Big[\int_0^t\Big\langle\gamma(x, \smooth{\gamma}\eta_s(x))
\Delta \big(f(\cdot)r(\cdot,\smooth{r}\eta_s(\cdot))\big)(x)
+F(x,\smooth{F}\eta_s(x)) f(x),
\eta_s(dx)\Big\rangle ds\Big].
\end{multline} 
We take a sequence of test functions $f_n$ increasing to
$\exp(\lambda x)$, with $\lambda>0$. Having passed
to the limit,
substituted in~(\ref{special case first moment equation}) and 
expanded the term involving the Laplacian, we obtain
\begin{multline}
\label{exponential moment}
    \IE[\langle \exp(\lambda x),\eta_t\rangle]
    =
    \IE[\langle \exp(\lambda x),\eta_0\rangle]
    +
    \IE\Big[\int_0^t\Big\langle\gamma(x, \smooth{\gamma}\eta_s(x))
    \exp(\lambda x)
    \Big\{
        \lambda^2 r(x, \smooth{r}\eta_s(x))
    \\ {}
        + 2\lambda\partial_x\big(r(\cdot,\smooth{r}\eta_s(\cdot)\big)(x)
        + \partial_{xx} r(\cdot,\smooth{r}\eta_s(\cdot)\big)(x)
    \Big\} 
    + F(x,\smooth{F}\eta_s(x)),
    \eta_s(dx)\Big\rangle ds\Big] .
\end{multline} 
Evidently, in much the same way as in the proof of
Lemma~\ref{lem:eta_f_bound},
our conditions on $r$ and $F$ ensure
that the integrand on the right hand side 
of~(\ref{exponential moment})
is bounded by $C\langle\exp(\lambda x),\eta_s(dx)\rangle$ for some constant
$C$. Gronwall's inequality then implies that 
$\IE[\langle\exp(\lambda x),\eta_t(dx)\rangle]$ 
is uniformly bounded for $t\in [0,T]$, and an application of Markov's inequality
tells us that for any $\delta >0$,
$\IP[\eta_t([R,\infty])>\delta]\to 0$ as $R\to\infty$. A symmetric argument
yields the corresponding bound for $\IP[\eta_t([-\infty, -R])>\delta$ and the 
result is proved.
\end{proof}

@@

\comment{TODO: check that $\eta_s^N$ having an $N$ or not is okay}

@@

\comment{perhaps explain: the next proposition is useful because we'll want to use weak convergence of $\smooth{\epsilon}\eta_s$; but in the martingale we have the term on the left in the proposition; probably there are other ways to do this.}
\comment{ACTUALLY BETTER we want to transfer $T_\epsilon^2$ over onto $\eta$; this happens in two steps, first this proposition; next using that $T_{\epsilon^2}f - f$ is small; ALISON}

@@

Combining the above, we have reduced the problem to showing that 
\begin{equation}
\label{target expression}
	\lim_{\epsilon\to 0}
	\int_0^t\big\langle \int |\rho_\epsilon*\eta_s(y)-\rho_\epsilon*\eta_s(w)|
\left(\rho_\epsilon*\eta_s(y))^k+\rho_\epsilon*\eta_s(w)^k\right)
p_{\epsilon^2}(w,y)dy, \eta_s(dw)\big\rangle ds
=0,
\end{equation}
\comment{check this: which we prove by showing the expectation of this positive quantity goes to zero; e.g., see Lemma 7.5(?): ALISON}.

@@

\comment{Probably should put a bit more detail in here?}
\comment{key bit is that we can always write $M_s(w)$ as a convolution of $p_{\epsilon^2/2}$ with something: ALISON}

@@

    The analogue of Lemma~\ref{lem:limit_mgale},
    in particular that the martingale properties are preserved under passage to the limit,
    are proved in Section~\ref{sec:lookdown_generator_proofs}.
    \comment{TODO: refer to correct thing or just say "like in 6.6": PETER}
    Finally, we must show that the limiting lookdown process $\lp$
    projects to the limiting process $\eta$, i.e., a solution of the martingale
    problem in Theorem~\ref{thm:nonlocal_convergence}.
    Let $N_k \to \infty$ be a sequence along which $\lp^{N_k}$ converges.
    By Theorem~\ref{thm:nonlocal_convergence},
    there is a subsequence $N_{k(j)}$ along which the projected population processes
    $\eta^{N_{k(j)}}$ converge, and the limit solves the martingale problem.
    Thus any limit point of $\lp^N$ projects to a population process $\eta$
    solving the martingale problem of Theorem~\ref{thm:nonlocal_convergence}.
\end{proof}

@@

We postpone the proof of this Lemma until we have shown how it yields to compact containment.
First, we show that this implies compact containment for arbitrary
compactly supported functions.

@@

In other words, the processes $\lp^N$ stay in 
the set
$$
    \left\{
        \lp \in \lpmeasures \;:\;
        \langle f_k, \lp \rangle \le C_k
        \text{ for all } k \ge 1
    \right\} ,
$$
for all $0 \le t \le T$
with uniformly high probability,
a set which is compact if we choose $\{f_k\}$ appropriately.
% \comment{
%     Note that we could instead of the vague topology on locally finite measures,
%     take the space of measures with $\langle h, \xi \rangle < \infty$
%     for which $\xi_n \to \xi$ if $\langle f, \xi \rangle < \infty$ for all $f$ for which $f \le c_f h$,
%     which is completely metrizable
%     as in Condition 2.1 of \citet{etheridge/kurtz:2019}.
% }
